Peer Assessment 1 - Reproducable Research
========================================================


## Load and process the data

The first step is to load the file `activity.csv` from the working directory.

```{r}
activity.data <- read.csv("activity.csv")
```


## What is mean total number of steps taken per day?

In order to aggregate the step count data into days, we will use functions from the `dplyr` package.

```{r}
library(dplyr)
activity.days <- activity.data %.% 
    group_by(date) %.% 
    summarise(steps =sum(steps, na.rm = TRUE))
```

Using this aggregated data, which creates the sum of steps for each day in the dataset, we can:

* Plot a histogram of the total number of steps taken each day  
* Calculate the median total steps taken each day
* Calculate the mean total steps taken each day

```{r fig.width=7, fig.height=6}
hist(activity.days$steps, main = "Histogram of total\nsteps taken each day", breaks = 10)

original.median <- median(activity.days$steps, na.rm = TRUE)
original.mean <- mean(activity.days$steps, na.rm = TRUE)


original.median <- format(original.median, nsmall = 2, big.mark = ",")
original.mean <- format(original.mean, nsmall = 2, big.mark = ",")
```

The calculated median of this data, ignoring `NA` values, is **`r original.median`**.

The calculated mean of this data, ignoring `NA` values, is **`r original.mean`**.

## What is the daily activity pattern?

We can also aggregate the data in 5-minute intervals so we can plot the average daily movement pattern.

```{r}

activity.interval <- activity.data %.% 
    group_by(interval) %.%
    summarise(steps = mean(steps, na.rm = TRUE))

plot(activity.interval$interval, activity.interval$steps, type = "l",
     main = "Average Steps by 5 Minute Interval",
     xlab="Interval",ylab="Average Number of Steps")
```

On average, the interval **`835`** contains the most steps on any given day.

```{r}
activity.interval$interval[activity.interval$steps == max(activity.interval$steps, na.rm = TRUE)]

```

## Imputing missing values

```{r}
na.idx <- is.na(activity.data$steps)

na.total <- sum(na.idx)
```

If we calculate total number of rows with `NA` values in the `steps` column, we can see that there are **`r na.total`** rows with missing data.

In order to impute the missing data, we will duplicate the dataset, then replace each `NA` value with the mean number of steps for that interval, using values the `activity.imputed` dataset.

```{r}
activity.imputed <- activity.data

for(i in seq_along(activity.imputed$steps)){
    if(is.na(activity.imputed$steps[i])){
        
        impute.data <- activity.interval[activity.interval$interval == 
                                             activity.imputed$interval[i],]
        
        activity.imputed$steps[i] <- impute.data$steps
    }
    
}
```

Aggregating the imputed data, we can plot a histogram of the total number of steps taken each day and calculate the median and mean for total steps taken in a day with this new data.

```{r fig.width=7, fig.height=6}
imputed.days <- activity.imputed %.% 
    group_by(date) %.%
    summarise(steps = sum(steps))

hist(imputed.days$steps, main = "Histogram of total (imputed)\nsteps taken each day", breaks = 10)

imputed.median <- median(imputed.days$steps, na.rm = TRUE)
imputed.mean <- mean(imputed.days$steps, na.rm = TRUE)

imputed.median <- format(imputed.median, nsmall = 2, big.mark = ",")
imputed.mean <- format(imputed.mean, nsmall = 2, big.mark = ",")

```

How do the new median and mean calculations differ from the origninal?

| Data Type | Median | Mean |
|:------:|------:|-----:|
|Original| `r original.median` | `r original.mean`|
|Imputed| `r imputed.median` | `r imputed.mean`|

Two trends stand out:

1. The original median and mean are both lower than that of the imputed dataset.
2. There is no difference between the imputed median and mean, but there is a difference between the original median and mean.

## Are there differences in activity patterns between weekdays and weekends?

First, we need to create a factor variable, `day.type` to flag if a day is a weekday or a weekend day. To accomplish this, we will use the `ymd()` function from the `lubridate` package to convert the date strings into a weekday same, such as Monday, Tuesday, Wednesday, etc. We will then run a `for` loop to create a `day.type` column that will indicate if a particular observation is on a weekend (Saturay or Sunday) or a weekday (all other days).

```{r}
library(lubridate)

activity.data <- activity.data %.% 
    mutate(weekday = weekdays(ymd(as.character(date))))

activity.data$day.type <- "Weekday"

for(j in seq_along(activity.data$weekday)){
    if(activity.data$weekday[j] == "Saturday"){
        activity.data$day.type[j] <- "Weekend"
    }
    if(activity.data$weekday[j] == "Sunday"){
        activity.data$day.type[j] <- "Weekend"
    }
}

activity.data$day.type <- as.factor(activity.data$day.type)
```

We can then plot the data to see how daily activity patterns weekdays differ from the pattern on weekends.

```{r}

activity.daytype <- activity.data %.% 
    group_by(day.type, interval) %.%
    summarise(steps = mean(steps, na.rm = TRUE))

library(ggplot2)

ggplot(activity.daytype, aes(x = interval, y = steps, group = day.type))+
    geom_line()+
    facet_grid(day.type~.)+
    labs(x = "Interval", y = "Average Number of Steps")+
    theme_bw()

```